Sustainable Pathways for Irish homes in changing TIMES

### 2021-07-05 manuscript.tex
Writeful Free - Done

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
Write after paper is completed

@@

Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results, so Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy and distinctive residential sector. While also examining how researchers have previously used Energy System Optimization Models (ESOMs) to analyse the residential sector.
Section \ref{LitR} describes the methodology used to develop the residential sector while also outlining the input data. The process of applying constraints, technologies, costs and demands in the model.  
Section \ref{Results} outlines the results obtained from selected scenarios. Section \ref{Discussion} discusses the strengths and weakness of the newly developed model and how it differs from other models. Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.

@@

Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.

\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

Leibowicz et al. developed an expanded a version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas. Austin was chosen because of the high granularity data availability \cite{EnergyInc.}. Leibowicz et al. results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ). However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock. Leibowicz et al. also finds electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings. These insights show the strength of an ESOM, but the not all costs are taken into account in Leibowicz study, the capital cost associated with more thermally efficient buildings are not included.

Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storages affect wind power investments in Denmark. The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland and Germany, in order to cover import/export. Exogenous user constraints applied to the model, based on current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030 and Individual oil boilers are phased out by 2030. The results by find when investments in individual heat pumps are allowed, air to water heat pumps are installed in non-district heating areas. This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments. The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies. Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.

Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island. 
The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach. Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy systems perspective, one reason for this is the wind power could also be stored in communal batteries. When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by a individual battery and only 36\% by a communal battery. While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights. The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.

@@

In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in transport, industry and services sector in Ireland. The bottom up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS. Three scenarios are explored, namely, the reference scenario, energy efficiency scenario and energy efficiency+ scenario. The reference assume that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020. In the reference scenario the residential total energy consumption rose 6.7 \% in 2020, compared to 2008. The share of space and water heating decreased from 82 \% to 77 \% in that period. In the energy efficiency scenario 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulted in a reduction in of energy consumption of 9.9 \% in 2020, compared to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. The results in a reduction in of energy consumption of 16.5 \% in 2020, compared to the reference scenario.Rogan et al. discussed the barriers, obstacles to investment, split incentives and information gaps which restrict energy efficiency improvements in the residential sector. The importance for policy to not only look at number of retrofit, but also the depth of retrofit works, is shown by the difference in scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, Thellufsen et al. \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland. A sensitivity analysis was preformed on investment costs, heat grid losses and discount rates, however in all cases the inclusion of district heating is optimal. The final results find district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal. Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector only models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland, some of which are outlined below.

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the residential sector, their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000 Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Mircsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the predicted scenario, energy savings easily outweighing programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation and drugs and mortality, were quantified, this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
into the computer model. the modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted, energy consumption predicted by the model is considerably higher than given in the national energy balance, due to intertal temperature assumptions. 
The model showed the energy-efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.

In 2011 Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model is predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined. Nor is the affect of internal temperatures on energy efficiency improvements.  
\par In 2013 Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The heat load for a statistical occupancy and year was established. The default values used in some of the BER database variables where highlighted. For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation. Ahern et al. found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset. The model thermal improvement measures resulted a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, this resulted in an average 12-year payback. Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.

\par In 2015 Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwelling, to examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model used, relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%. DEAP is the national tool for BER calculation and checking building regulation compliance. The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied. BER database filters were used in this model, the BER filters from this study, were updated and applied in TIM. A sensitivity analysis is performed on the effects of internal temperature and rebound. It was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when compared the baseline scenario. This shows the the type of retrofit and type of house can have a huge affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.

@@

The listed models are some of models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.

The models described in this section included a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery systems location, evaluating health effects of retrofitting and integration of bio-energy and heat. The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the every growing BER database is providing high data granularity input to these models. The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at specific policy. The complexity of the models have increased significantly in the last 20 years, in Ireland the models starting out in Microsoft Excel and switching to Microsoft SQL. 
There has been few studies on the Irish residential sector using ESOMs, as the residential bottom-up models are preferred when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation or transport. The use of ESOM to explore DH in Ireland, allows for insights which the bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions. GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. While, gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. 
The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat of among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}. Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.

@@

In Ireland the decarbonisation of the electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. 

>>>>

\begin{abstract}
%% Text of abstract
%Ireland's residential sector has particularly underachieved in previous climate targets. Located on the peripherals of Europe, with a high share of low thermal efficient detached dwellings with carbon intensive heating fuels, provides some distinctive barriers to decarbonising Ireland's residential sector. The disadvantageous starting position has impeded residential decarbonisation progress thus far. \par
%Energy system optimization models (ESOMs) have been used extensively to inform pathways in addressing long-term energy challenges which provides insights to decision makers on issues related to climate and energy policy. \par
%The TIMES-Ireland Model (TIM) is a newly developed optimisation model of the Irish energy system, which calculates the  cost-optimal decarbonisation pathway to meet future energy service demands while achieving Ireland's legally binding 2030 and 2050 climate targets.
Write after the paper is completed

@@

Understanding residential energy consumption in Ireland is key to exploring future residential decarbonisation pathways, and interpreting the results, so Section \ref{sec:Introduction} of this paper looks at Ireland's climate targets, energy policy and distinctive residential sector. While also examining how researchers have previously used Energy System Optimization Models (ESOMs) to analyse the residential sector.
Section \ref{LitR} describes the methodology used to develop the residential sector while also outlining the input data. The process of applying constraints, technologies, costs, and demands in the model.  
Section \ref{Results} outlines the results obtained from the selected scenarios. Section \ref{Discussion} discusses the strengths and weakness of the newly developed model and how it differs from other models. Lastly, section \ref{Conclusion} draws some conclusions on the model, results and discussion on future work.

@@

Energy System Optimisation Models (ESOMs) can be used to explore the least cost pathways, as they consider a large number of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM covers all energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section.

\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) for TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

Leibowicz et al. developed an expanded a version of the Open Source Energy Modeling System (OSeMOSYS) \cite{Howells2011OSeMOSYS:Development.} that integrates electricity generation and storage, residential building energy service demands, and end-use technologies \cite{Leibowicz2018OptimalServices}.Endogenous constraints on the market penetration rate of each technology and hourly service demand profiles are established and applied to the residential sector of Austin, Texas. Austin was chosen because of the high granularity data availability \cite{EnergyInc.}. Leibowicz et al. results find net present costs in the \textit{``Policy scenarios"} ( emissions declining linearly toward 10\% of their current level by 2050) are higher than \textit{``No Policy scenarios"} (No $CO_2$ regulations imposed ). However, the percentage increase in net present cost due to the policy diminishes with a more thermally efficient building stock. Leibowicz et al. also found electrification to be the dominant strategy for cost-effectively decarbonizing urban residential dwellings. These insights show the strength of an ESOM, but not all costs are taken into account in Leibowicz study, the capital costs associated with more thermally efficient buildings are not included.

Hedegard and Münster \cite{Hedegaard2013InfluenceOperation} utilised  Balmorel model \cite{Wiese2018BalmorelModel} to handle the integrated power and district heating system, to analyse how heat pumps with complementing heat storage affect wind power investments in Denmark. The model was disaggreated between a western and eastern Denmark region and the model covers Norway, Sweden, Finland, and Germany, in order to cover import/export. Exogenous user constraints applied to the model, based on the current Danish policy includes: Wind power must provide at least 50\% of the national electricity consumption by 2020, coal is phased out in heat and power plants by 2030, and individual oil boilers are phased out by 2030. The results find when investments in individual heat pumps are allowed, air-to-water heat pumps are installed in non-district heating areas. This represents a significant electricity demand in identifying individual heat pumps as highly competitive and can contribute significantly to facilitating larger wind power investments and reducing system costs, fuel consumption, and $CO_2$ emissions.
Hedegard and Münster found the system benefits of adding individual heat storage to the heat pumps are moderate, the main benefit is a reduced need for peak/reserve capacity investments. The Balmorel model has provided insights, but it does not cover all aspects of the power and heat system, such as start-up costs, minimum load requirements, or part load efficiencies. Nevertheless, the insights are valuable and the results can be soft-linked to other models for deeper analysis.

Marczinkowski and Østergaard \cite{Marczinkowski2018ResidentialSystems} use EnergyPLAN model \cite{stergaard2015ReviewingSimulations} to explore the feasibility of photovoltaic (PV) systems installed at the household level in combination with batteries in Samsø, which is known for being Denmark’s renewable energy island. 
The two approaches are observed for the location of the batteries in the PV system; communal battery system or an individual residential battery approach. Marczinkowski and Østergaard found communal batteries slightly more favourable from a technical energy system perspective, one reason for this is the wind power could also be stored in communal batteries. When identical communal and individual battery capacities where compared, it was fond  64\% of residential electricity demand is satisfied by an individual battery and only 36\% by a communal battery. While the battery location was explored, the PV system location was not, further investigation into the optimal PV system location could provide further insights. The household occupancy behaviour with and without household batteries on the PV system could alter the optimal pathway.

@@

In 2014, Rogan et al. \cite{Rogan2014LEAPsSystem} linked OSeMOYOS, an ESOM to the Long Range Energy Alternatives Planning System (LEAP) simulation model, to explore specific energy efficiency policies not only in the residential sector, but also in transport, industry and services sectors in Ireland. The bottom-up development of the residential sector in LEAP was combined with the cost optimisation capabilities of OSeMOYOS. Three scenarios are explored, namely, the reference scenario, energy efficiency scenario and energy efficiency+ scenario. The reference assume that 300,000 existing dwellings will undergo a shallow retrofit from 2009 to 2020. In the reference scenario the residential total energy consumption rose 6.7 \% in 2020, compared to 2008. The share of space and water heating decreased from 82 \% to 77 \% in that period. In the energy efficiency scenario, 800,000 residential dwellings are retrofitted between 2009 and 2020, combined with the effect of other policies such as building regulations and lighting, resulting in a 9.9 \% reduction of energy consumption in 2020, compared to the reference scenario. The energy efficiency+ scenario also allows for 800,000 residential dwellings to be retrofitting, but accounts for deeper retrofits. The results in a reduction in of energy consumption of 16.5 \% in 2020, compared to the reference scenario.Rogan et al. discussed the barriers, obstacles to investment, split incentives, and information gaps which restrict energy efficiency improvements in the residential sector. The importance for the policy to not only look at the number of retrofits, but also the depth of retrofit work, is shown by the difference in scenarios. The model also found considerable scope for energy savings to be made through improved building regulations.

In 2018, Thellufsen et al. \cite{Thellufsen2019ImplementingIreland} investigated the potential for the implementation of district heating in both the commercial and residential sectors of Ireland. The paper analyses the $CO_2-80$ scenario ( 80 \% $CO_2$ reduction in 2050, compared to 1990 ) with and without district heating using the previous Irish TIMES model and EnergyPLAN. The previous version of Irish TIMES did not account for district heating, so EnergyPLAN is used to investigate the operation and  feasibility of district heating in Ireland. A sensitivity analysis was preformed on investment costs, heat grid losses, and discount rates, however in all cases the inclusion of district heating is optimal. The final results find that district heating reduces fuel consumption by 3.5 \% and the conversion of 37 \% of the Irish heat demand to district heating is found to be optimal. Both the TIMES model and climate target are superseded and the study did not include the future heat savings expected from retrofitting, nevertheless this study has provided insights on the potential of district heating.

\par  The majority of previous modelling on the residential sector in Ireland has involved bottom-up residential sector models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on the input data and methodology approach used in Ireland, some of which are outlined below.

Clinch and Healy analysed different aspects of thermal comfort in the Irish residential sector. This was the first comprehensive analysis of the residential sector, their studies examined fuel poverty \cite{Healy2002FuelIreland}, excess winter mortality \cite{Clinch2000HousingMortality}, market failures \cite{Clinch2000DomesticFailure} and energy efficiency  \cite{Clinch2003ValuingModel,Clinch2000DomesticFailure,Clinch2001Cost-benefitEfficiency,Clinch2001ModellingEfficiency}. In 2000, Clinch and Healy developed and published a domestic energy efficiency model to analyse the cost-benefit of energy efficiency \cite{Clinch2001Cost-benefitEfficiency}.The Energy-Assessment Model (EAM) is a bottom-up model developed in Microsoft Excel which incorporates 1,824 representative dwellings ( 8 building archetypes, 12 levels of insulation and 19 heating systems ), this was not linked to other sectors of the economy and only looked at space and water heating. A cost-benefit model (CBM) was fitted to the EAM to facilitate the conversion of the physical estimates into monetary amounts. Under the predicted scenario, energy savings easily outweighing programme costs, representing a benefit-cost ratio of 1.7.The benefits of reduced sickness, hospitalisation and drugs and mortality, were quantified, this moved the  benefit-cost ratio up to 3.0.
The same model was used by Clinch, Healy and King \cite{Clinch2001ModellingEfficiency} in addition three models of householder spending on energy are developed here to to evaluate the human behavioural component
into the computer model. The modelling process involved developing a computer model of the energy performance of the existing dwelling stock. It was noted that energy consumption predicted by the model is considerably higher than given in the national energy balance, due to internal temperature assumptions. 
The model showed that the energy efficiency programme ( all houses retrofitted to 1997 thermal building standards in 10 years ) represents a saving of 24 \% when compared to the reference scenarios.

In 2011, Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on the energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined, nor is the internal temperatures affect on energy efficiency improvements.  
\par In 2013, Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The heat load for a statistical occupancy and year was established. The default values used in some of the BER database variables where highlighted. For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs despite a 2001 survey found 18\% of detached housing in Ireland had no roof insulation. Ahern et al. found dwellings built post-2000 consume the greatest amount of energy (kWh/annum), due to their large floor area. Concluding that the majority of energy efficiency progress for thermal use in Irish dwellings is being offset. The model thermal improvement measures resulted a reduction of running costs and CO2 emissions by an average of 65\% for houses constructed prior to 1979, and this resulted in an average 12-year payback. Houses built in 1979 or after resulted in an average 26\% reduction in running costs and $CO_2$ emissions from retrofit measures.

\par In 2015, Dineen et al.\cite{Dineen2015ImprovedSource} built on previous work in examining new dwellings, by examining energy savings potential of energy efficiency improvement measures of the existing dwelling stock. The Dwelling Energy Assessment Procedure (DEAP) model used, relied on the BER database, which accounted for 18\% of all Irish dwellings, currently the database has more than doubled and accounts for over 50\%. DEAP is the national tool for BER calculation and checking building regulation compliance. The model used 175 archetype dwellings ( 5 building types, 7 energy bands and 5 wall construction types ) and 6 different retrofit measures were applied. BER database filters were used in this model, the BER filters from this study were updated and applied in TIM. A sensitivity analysis is performed and it was found that there was approximately a 7.5\% change in energy consumption for every 1$^{\circ}$C internal temperature shift. Concluding, BER standard assumptions overestimate dwelling energy consumption and that the average internal temperature across the Irish dwelling stock may be closer to 18$^{\circ}$C than 21$^{\circ}$C.

The Archetype Dwelling Energy Model (ArDEM-SQL) was developed by Mac Uidhir et al.\cite{Uidhir2020ImprovingDecisions}. ArDEM-SQL converts the Irish BER public database into Microsoft SQL language and then calculates annual energy consumption to simulate the potential for improved energy efficiency gains within the existing retrofit programs. Mac Uidhir et al. specifically examined the Better Energy Homes (BEH) scheme. The five most common retrofit combinations are simulated for each of the 9 archetypes ( 3 energy rating types and 3 building types ). The results show that the alternative retrofit scenarios can achieve additional energy efficiency gains of up to 86\% can be achieved when comparing the baseline scenario. This shows the the type of retrofit and type of house can have a huge affect on the energy saved. 
Greater energy savings are possible when retrofit measures are applied to less energy efficient dwellings (E, F, and G-ratings ), albeit the ``rebound effect" was not accounted for due to insufficient data on the scale of the rebound effect and the relative difference between the two scenarios examined is considered negligible.

@@

The listed models are some of the models used within SEAI's NEMF to model the impact of incentives and opportunities created by specific policy instruments.

The models described in this section included a range of national ESOMs and bottom-up residential models. The models have varied objectives such as exploring consumer choices, PV battery system location, evaluating health effects of retrofitting, and integration of bioenergy and heat. The thermal efficiency of dwellings is a common objective in Irish models, due to the poor thermal efficiency of the dwelling stock and the every growing BER database is providing high data granularity input to these models. The internal temperature is also analysed in some of the studies, a recurring theme is the wide range of internal temperatures. 
The bottom-up residential models are typically simulation models looking at a specific policy. The complexity of the models has increased significantly in the last 20 years, in Ireland, the models starting out in Microsoft Excel and switching to Microsoft SQL. 
There has been few studies on the Irish residential sector using ESOMs, as the residential bottom-up models are preferred when focusing on that sector. The studies on Irish ESOM typically do not focus on the residential sector, instead they focus on the overall energy system, electricity generation, or transport. The use of ESOM to explore DH in Ireland, allows for insights which bottom-up simulation models cannot provide. The connection between waste heat in electricity and industry can show the optimal solution for district heating in commercial and residential sectors.

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings, and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil, accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential Greenhouse Gas (GHG) emissions. GHG emissions from coal in the residential sector also declined to 6.9\% in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. While, gas availability is limited to some urban areas, and accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. 
The residential sector relied on fossil fuels for over 90\% of thermal energy in 2018, consequently Ireland currently has the lowest renewable heat among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}. Preliminary results from the Environmental Protection Agency (EPA) have indicated a 9\% increase of residential GHG emissions in 2020, mainly due to low home heating oil prices and covid-related restrictions.

@@

In Ireland, the decarbonisation of electricity is underway, with carbon intensity falling annually to 375 $gCO_2/kWh$ in 2018, which is 40\% lower than 2005 levels \cite{Howley2005ENERGY-RELATEDIreland}. 
